---
title: "Analysis of duration of response in oncology studies"
author: Lu Mao, Lu Tian, Bo Huang, Lee-Jen Wei
format: 
  html:
    html-math-method: katex
    toc: true
    toc-depth: 4
    toc-expand: true
execute: 
  # eval: false
  echo: true
  cache: true
bibliography: references.bib
---

This R-code implements the analysis of duration-of-response data described in @mao:2024 for oncology studies .

### Background

Duration-of-response (DOR) is often utilized to assess efficacy of oncology therapies. While valid procedures are available for estimating the mean [@huang2018; @huang2020; @huang2022], it is unclear how to estimate the distribution of DOR, which may provide a fuller picture. Simple Kaplan–Meier estimates based on the responders can result in distorted conclusions due to the existence of non-responders (whose DOR is zero).

### Notation and Methodology

#### Full and Observed Data

For a responder, let $T_1$ denote the time to response. Let $T_2$ denote the time to disease progression or death. If a patient experiences progression/death first, then $T_1=\infty$. The DOR can be defined as $[T_2-T_1]_+$, where $x_+=xI(x>0).$ Thus, DOR is $T_2-T_1$ for responders, i.e., those with $T_1<\infty$, and 0 for nonresponders. For notational convenience, set $T_1=T_2$ for nonresponders.

Due to limited follow-up, the entire distribution of DOR is not identifiable. Instead, we define the restricted DOR within a time window as $D=T_2\wedge \tau -T_1\wedge \tau$, where $\tau$ is a pre-specified restriction time. Our goal is to estimate the survival function of $D:$ $$S_D(t)=P(D>t), \,\,\,\,\, t\in [0, \tau].$$

Let $C$ denote the censoring time independent of $(T_1, T_2)$ and satisfying $P(C\geq \tau)>0$. Then we observe $(X_1,\Delta_1,X_2,\Delta_2)$, where $X_1=\min(T_1\wedge\tau, C)$, $\Delta_1=I(T_1\wedge\tau\leq C)$, $X_2=\min(T_2\wedge\tau,C)$, and $\Delta_2=I(T_2\wedge\tau\leq C).$ Write $\tilde D=X_2 - X_1$. Here both $T_1$ and $T_2$ are subject to the same censoring time $C$. The observed data consist of $n$ independent and identically distributed (i.i.d.) copies of $(X_1,\Delta_1,X_2,\Delta_2)$: $\{(X_{1i},\Delta_{1i},X_{2i},\Delta_{2i}): i=1,\ldots,n\}.$ Write $\tilde D_i=X_{2i}- X_{1i}$.

#### Estimation and Inference

A simple estimator of $S_D(t)$ is the inverse probability censoring weighted (IPCW) estimator: $$
\hat S_D(t)=n^{-1}\sum_{i=1}^n  \frac{\Delta_{2i}}{\hat{G}_C(X_{2i})} I(\tilde D_i> t),
$$ where $\hat{G}_C(t)$ is the Kaplan--Meier estimator of $G_C(t)=P(C>t)$ based on $\{(X_{2i}, 1-\Delta_{2i}), i=1, \cdots, n\}$. To see $\hat S_D(t)$ as a valid estimator, note that $\Delta_{2}=1$ implies that $X_2=T_2\wedge\tau$, $X_1=T_1\wedge\tau$ and thus that $\tilde D = D$. Hence, $E\left\{\Delta_2I(\tilde D>t)/G_C(X_2)\right\}= E\left\{\Delta_2I(D>t)/G_C(T_2\wedge\tau)\right\} =E\left[E\left\{\Delta_2I(D>t)/G_C(T_2\wedge\tau)\mid T_1, T_2\right\}\right] =E\{I(D>t)\}=S_D(t)$, where the second equality follows by the independence of $C$ and $(T_1, T_2)$. Moreover, we can show that $\hat S_D(t)$ converges weakly to a Gaussian process with an easily estimable variance function.

The median DOR can thus be estimated by $\inf\{t: \hat S_D(t)\geq 0.5\}$, whose variance can be estimated by the perturbing the influence functions of $\hat S_D(t)$ using standard normal random noises.

### Usage and Example

#### R-program

To use the program, download `DORfunctions.R` from repo <https://github.com/lmaowisc/dor>. The R-file contains all functions needed to perform the analysis. We also need the standard `survival` package.

```{r}
library(survival)
source("DORfunctions.R")
```

The main function is `dorfit(x1, delta1, x2, delta2, tau, med_inf = FALSE)`.

**Input**

-   `x1` time to earliest of response, outcome event, or censoring ($X_1$)
-   `delta1` indicator of response ($\Delta_1$)
-   `x2` time to earlier of outcome event and censoring ($X_2$)
-   `delta2` indicator of outcome event ($\Delta_2$)
-   `tau` restriction time ($\tau$)
-   `med_inf` whether to make inference on the median DOR (default is `FALSE`)

**Output**

-   `t1` vector of times ($t$)
-   `surv1` IPCW estimates of survival function $\hat S_D(t)$
-   `se1` pointwise standard error of $\hat S_D(t)$
-   `med` median estimate of DOR
-   `med_se, med_lo, med_hi` these are standard error, lower and upper limits of the 95% confidence interval of the median, respectively, if `med_inf = TRUE`; all `NULL` if `med_inf = FALSE`

#### A simulated example

Let's simulate a data set of size $n=200$.

```{r}
# set seed for random number generation
set.seed(2023)

## simulate data
n <- 200
# set restriction time
tau <- 1.25

# full data
t1 <- pmin(rexp(n), runif(n, 0, 0.75)) # response time
t2 <- rexp(n)*1.5 # outcome event time
t1[t1>t2] <- t2[t1>t2] # set t1 = t2 for non-responders
c <- runif(n, 0, 1.5) # censoring time

# observed (censored) data
x1 <- pmin(t1, c)
delta1 <- 1*(t1<c)
x2 <- pmin(t2, c)
delta2 <- 1*(t2<c)
```

So the observed data consist of `x1`, `delta1`, `x2`, and `delta2`. Under restriction time `tau`$= 1.25$. We analyze the DOR distribution using the main function `dorfit()`, with inference on the median.

```{r}
fitLM <- dorfit(x1, delta1, x2, delta2, tau, med_inf = TRUE)
```

Let's plot the estimated $S_D(t)$ with 95% pointwise confidence limits.

```{r}

za <- qnorm(0.975)
## marginal survival function P(D > t)
t <- fitLM$t1 # times
St <- fitLM$surv1 # P(D > t)
se <- fitLM$se1 # se of P(D > t)

# set up the figure
plot(c(0, 1.2), c(0, 1), type="n",xlab="t",ylab=expression(S[D](t)))

# plot the marginal curve P(D > t)
lines(stepfun(t, c(1,St)),lwd=2, do.points = FALSE) # marginal
lines(stepfun(t, c(1,St + za * se)),lty = 3, do.points = FALSE) # upper 95% CI 
lines(stepfun(t, c(1,St - za * se)),lty = 3, do.points = FALSE) # lower 95% CI 

```

We can see that the median is close to 0.6. The precise number and confidence interval (CI) can be obtained from `fitLM`.

```{r}
#| output: false

######## get the median ####
med <- fitLM$med
med
# [1] 0.551916

med_lo <- fitLM$med_lo
med_hi <- fitLM$med_hi
med_lo
# [1] 0.4018724
med_hi
# [1] 0.7019595
```

So the estimated median DOR (95% CI) is `r round(med, 3)` (`r round(med_lo, 3)`, `r round(med_hi, 3)`).
